---
title: "Exploratory Analysis of the ToothGrowth dataset"
author: "Graham King"
date: "24 January 2015"
output:
  html_document:
    css: report.css
---
```{r opts, echo=FALSE}
library(knitr)
opts_chunk$set(comment = NA, fig.width = 12, fig.height = 6, echo = FALSE, size = 'small')

read_chunk("ggInit.R")
read_chunk("tgIntro.R")
read_chunk("tgClean.R")
read_chunk("tgExplore1.R")
read_chunk("tgExplore3.R")
read_chunk("tgConfInt.R")
```
###Overview
This is an analysis of the ToothGrowth dataset from the R Datasets package.

The ToothGrowth help page has the following description of the dataset:

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

### Data cleaning
The dataset documentation suggests that this dataset should consist of 60 observations (10 subjects * 3 doses * 2 delivery methods)

```{r intro}
<<ggInit>>
<<tgIntro>>
```

From this I can see that tooth length is a continuous variable named `len`; delivery method is a factor variable named `supp` with levels `OJ` and `VC`; and the dose level is a numeric variable named `dose`. Given that dose represents three dose levels, this would be better as a factor.

The `supp` variable represents two delivery methods of Vitamin C, being orange juice and asorbic acid. However the levels are named 'OJ' and 'VC'. Presumably VC represents Ascorbic Acid. For clarity, I will rename the levels in this variable.
```{r clean, echo = TRUE}
<<tgClean>>
```

A check on the number of nulls in the dataset shows that there are none:
```{r intro2, echo = TRUE}
<<tgIntro2>>
```

### Exploratory Plots
First I plot the distribution of the tooth length variable by supplement:
```{r tgExplore1}
<<tgExplore1>>
```

Figure 1 shows that the mean tooth length is greater with the Orange Juice supplement than with the Ascorbic Acid. However, despite having a lower mean, the Ascorbic Acid sample has a greater variance, demonstrated by a larger standard distribution.

Having looked at how the supplement affects the tooth length distribution, I will now look at the effect that the supplement dose has.
```{r explore3}
<<tgExplore3>>
```

Figure 2 shows that a higher dose of supplement results in a longer tooth length.

The distributions are all approximately normal (i.e. identically distributed) although with a small amount of skew in some of the subsets of the data. I will assume that this is not enought to invalidate the use of *t* interval. I also assume that the values sampled are independent. This also means that I assume that the data is not paired, an assumption that is supported by the lack of subject identifier in the dataset.

### Confidence Intervals
```{r cifunction}
<<cifunction>>
```
The mean tooth length for each supplement type will fall between below lower and upper limits in 95% of samples of a similar size.

#### Confidence Interval 1
```{r ci1}
<<ci1>>
```

The mean tooth length for each dose will fall between below lower and upper limits in 95% of samples of a similar size:

#### Confidence Interval 2
```{r ci2}
<<ci2>>
```

The following t confidence interval shows the that the mean increase in tooth length length using supplement Orange Juice rather than Ascorbic Acid will be within the given range in 95% of samples of a similar size:

#### Confidence Interval 3
```{r ci3}
<<ci3>>
```
This small increase is also reflected in the close means and large overlap of the subsets of data seen in Figure 1. A greater difference can be seen by looking at the smaller dose against the larger dose. The following t confidence interval shows that the mean increase in tooth length with a 2.0mg dose rather than a 0.5mg dose will be within the given range in 95% of samples of a similar size:

#### Confidence Interval 4
```{r ci4}
<<ci4>>
```
This is also reflected in the difference between the means of 0.5mg and 2.0mg in Figure 2.

However, the greatest difference can be seen when comparing 0.5mg of the least effective supplement (Ascorbic Acid) with 2.0mg of Orange Juice. The following t confidence interval shows that the mean increase in tooth length with a 2.0mg dose of Orange Juice rather than a 0.5mg dose of Ascorbic Acide will be between the given range in 95% of samples of a similar size.:

#### Confidence Interval 5
```{r ci5}
<<ci5>>
```

****

### Appendix
Initialise ggplot parameters:
```{r ap1, eval = FALSE, echo = TRUE}
<<ggInit>>
```
Introduction:
```{r ap2, eval = FALSE, echo = TRUE}
<<tgIntro>>
```
Clean data:
```{r ap3, eval = FALSE, echo = TRUE}
<<tgClean>>
```
Figure 1:
```{r ap4, eval = FALSE, echo = TRUE}
<<tgExplore1>>
```
Figure 2:
```{r ap5, eval = FALSE, echo = TRUE}
<<tgExplore3>>
```
Confidence Interval 1:
```{r ap7, eval = FALSE, echo = TRUE}
<<ci1>>
```
Confidence Interval 2:
```{r ap8, eval = FALSE, echo = TRUE}
<<ci2>>
```
```{r ap8a, eval = FALSE, echo = TRUE}
<<cifunction>>
```
Confidence Interval 3:
```{r ap9, eval = FALSE, echo = TRUE}
<<ci3>>
```
Confidence Interval 4:
```{r ap10, eval = FALSE, echo = TRUE}
<<ci4>>
```
Confidence Interval 5:
```{r ap11, eval = FALSE, echo = TRUE}
<<ci5>>
```