---
title: "Analysis of the Statistical Properties of the Exponential Distribution"
author: "Graham King"
date: "24 January 2015"
output:
  html_document:
    css: report.css
---
```{r opts, echo = FALSE}
library(knitr)
library(markdown)
opts_chunk$set(comment = NA, fig.width = 12, fig.height = 8)
options(scipen = 999)

read_chunk("ggInit.R")
read_chunk("intro.R")
read_chunk("expSample.R")
read_chunk("expSampleMeans.R")
```

### Overview
The objective of this analysis is to look at the mean, variance and standard error of a simulation of numbers drawn at random from the exponential distribution. This distribution has a number of known properties of the theoretical total population, and this analysis will show that the above statistical properties of the sample dataset are close to the expected values as predicted by the theoretical population's properties.

### Introduction
```{r ggInit, echo = FALSE}
<<ggInit>>
```
```{r intro, echo = FALSE, eval = TRUE}
<<intro>>
```
The exponential distribution is defined by a 'rate parameter', known as 'lambda' (<math display="inline"><mo>&lambda;</mo></math>). It has a mean of <math display="inline"><mfrac bevelled="true"><mrow><mn>1</mn></mrow><mrow><mo>&lambda;</mo></mrow></mfrac></math> and its standard deviation is also <math display="inline"><mrow><mfrac bevelled="true"><mrow><mn>1</mn></mrow><mrow><mo>&lambda;</mo></mrow></mfrac></math>, meaning its variance is <math display="inline"><msup><mfenced><mfrac bevelled="true"><mrow><mn>1</mn></mrow><mrow><mo>&lambda;</mo></mrow></mfrac></mfenced><mn>2</mn></msup></math>. Finally its standard error of the mean, or 'SEM', is <math display="inline"><mfrac bevelled="true"><mrow><mo>&sigma;</mo></mrow><mrow><msqrt><mi>n</mi></msqrt></mrow></mfrac></math>, meaning that a distribution of the means of a number of samples of size *n* should have this standard distribution.

Below is a table of the variables used in this analysis, with their associated mathematical symbols, a description of what it represents and how they are calculated.:

Variable  |  Symbol | Description | Calculated As
--------- | ------- | ----------- | -------------
lambda | <math display="inline"><mo>&lambda;</mo></math> | rate parameter of the exponential distribution |  This analysis will use a lambda value of `r lambda`
mu | <math display="inline"><mo>&mu;</mo></math> | population mean | <math display="inline"><mfrac bevelled="true"><mrow><mn>1</mn></mrow><mrow><mo>&lambda;</mo></mrow></mfrac></math>
sigma | <math display="inline"><mo>&sigma;</mo></math> | population standard deviation | <math display="inline"><mfrac bevelled="true"><mrow><mn>1</mn></mrow><mrow><mo>&lambda;</mo></mrow></mfrac></math>
sigma2 | <math display="inline"><msup><mo>&sigma;</mo><mn mathsize="small">2</mn></msup></math> | population variance | <math display="inline"><msup><mfenced><mfrac bevelled="true"><mrow><mn>1</mn></mrow><mrow><mo>&lambda;</mo></mrow></mfrac></mfenced><mn>2</mn></msup></math>
sem | <math display="inline"><msub><mo>&sigma;</mo><mover><mi>x</mi><mo>&OverBar;</mo></mover></msub></math> | standard error of the mean (SEM) | <math display="inline"><mfrac bevelled="true"><mrow><mo>&sigma;</mo></mrow><mrow><msqrt><mi>n</mi></msqrt></mrow></mfrac></math>
xbar | <math display="inline"><mover><mi>x</mi><mo>&OverBar;</mo></mover></math> | sample mean | <math display="inline"><mfrac><mrow><mn>1</mn></mrow><mrow><mi>n</mi></mrow></mfrac><mrow><munderover><mo>&Sigma;</mo><mrow><mi>i</mi><mo>=</mo><mn>1</mn></mrow><mi>n</mi></munderover><msub><mi>x</mi><mi>i</mi></msub></mrow><mspace linebreak="newline" /><mtext mathsize="small">(where x is the sampled values)</mtext></math>
s2 | <math display="inline"><msup><mi>s</mi><mn mathsize="small">2</mn></msup></math> | sample variance | <math display="inline"><mfrac><mrow><mn>1</mn></mrow><mrow><mi>n</mi><mo>-</mo><mn>1</mn></mrow></mfrac><mrow><munderover><mo>&Sigma;</mo><mrow><mi>i</mi><mo>=</mo><mn>1</mn></mrow><mi>n</mi></munderover><msup><mfenced><mrow><msub><mi>x</mi><mi>i</mi></msub><mo>-</mo><mover><mi>x</mi><mo>&OverBar;</mo></mover></mrow></mfenced><mn>2</mn></msup></mrow></math>
s | <math display="inline"><mi>s</mi></math> | sample standard deviation | <math display="inline"><msqrt><msup><mi>s</mi><mn mathsize="small">2</mn></msup></msqrt></math>
muxbar | <math display="inline"><msub><mo>&mu;</mo><mover><mi>x</mi><mo>&OverBar;</mo></mover></msub></math> | mean of sample means | <math display="inline"><mfrac><mrow><mn>1</mn></mrow><mrow><mi>s</mi></mrow></mfrac><mrow><munderover><mo>&Sigma;</mo><mrow><mi>i</mi><mo>=</mo><mn>1</mn></mrow><mi>s</mi></munderover><msub><mover><mi>x</mi><mo>&OverBar;</mo></mover><mi>i</mi></msub></mrow><mspace linebreak="newline" /><mtext mathsize="small">(where s is the number of simulations)</mtext></math>
s2xbar | <math display="inline"><msubsup><mi>s</mi><mover><mi>x</mi><mo>&OverBar;</mo></mover><mn mathsize="small">2</mn></msubsup></math> | variance of sample means | <math display="inline"><mfrac><mrow><mn>1</mn></mrow><mrow><mi>s</mi><mo>-</mo><mn>1</mn></mrow></mfrac><mrow><munderover><mo>&Sigma;</mo><mrow><mi>i</mi><mo>=</mo><mn>1</mn></mrow><mi>s</mi></munderover><msup><mfenced><mrow><msub><mover><mi>x</mi><mo>&OverBar;</mo></mover><mi>i</mi></msub><mo>-</mo><msub><mo>&mu;</mo><mover><mi>x</mi><mo>&OverBar;</mo></mover></msub></mrow></mfenced><mn>2</mn></msup></mrow><mspace linebreak="newline" /><mtext mathsize="small">(where s is the number of simulations)</mtext></math>
sxbar | <math display="inline"><msub><mi>s</mi><mover><mi>x</mi><mo>&OverBar;</mo></mover></msub></math> | standard deviation of sample means | <math display="inline"><msqrt><msubsup><mi>s</mi><mover><mi>x</mi><mo>&OverBar;</mo></mover><mn mathsize="small">2</mn></msubsup></msqrt></math>

With lambda of `r lambda`, the distribution's theoretical statistics are:

Stat | Description | Value
---- | ----------- | -----
lambda | rate parameter | `r lambda`
mu | mean | `r mu`
sigma | standard deviation | `r sigma`
sigma2 | variance | `r sigma2`
n | number of samples | `r n`
sem | standard error of the mean | `r sem`

### Distribution of Sample
Here I take a random sample of `r n * numsims` independent variables from the exponential distribution and calculate their mean, variance and standard deviation.
```{r expSampleCalc, echo = FALSE, eval = TRUE}
<<expSampleCalc>>
```
Stat | Description | Value
---- | ----------- | -----
xbar | mean of sample | `r xbar`
s2 | variance of sample | `r s2`
s | standard deviation of sample | `r s`

These same sampled values will be split later into `r numsims` groups of `r n` for multiple sample means to be taken, but here I plot the distribution of all of the sampled values along with the sample mean and variance and the population mean and variance.
```{r expSamplePlot, echo = FALSE, eval = TRUE}
<<expSamplePlot>>
```

Figure 1 demonstrates that the mean of this sample (`r xbar`) is very close to the theoretical mean of the whole population (`r mu`) and that the sample variance (`r s2`) is similarly very close to the population variance (`r sigma2`). The distribution of these random variables reflects very clearly the exponential distribution, as shown by overlaying the true exponential distribution density, <math display="inline"><mi>exp</mi><mfenced><mo>&lambda;</mo></mfenced></math>.

### Distribution of Sample Means
I next split this sample of `r 40000` into `r numsims` groups of `r n` and take the mean of each sample. From this list of means I can calculate its mean (muxbar), variance and standard devitaion.
```{r expSampleMeansCalc, echo = FALSE, eval = TRUE}
<<expSampleMeansCalc>>
```
Stat | Description | Value
---- | ----------- | -----
muxbar | mean of sample means | `r muxbar`
s2xbar | variance of sample means | `r s2xbar`
sxbar | standard deviation of sample means | `r sxbar`

I then plot the distribution of these sample means together with their mean and standard deviation.
```{r expSampleMeansPlot, echo = FALSE, eval = TRUE}
<<expSampleMeansPlot>>
```

Whereas the sample of random exponential values followed the natural exponential distribution, Figure 2 shows that these sample means follow a normal distribution. This is shown clearly by overlaying normal distribution <math display="inline"><mi>N</mi><mfenced><mo>&mu;</mo><mfrac bevelled="true"><mrow><msup><mo>&sigma;</mo><mn>2</mn></msup></mrow><mrow><mi>n</mi></mrow></mfrac></mfenced></math>; that is to say, the normal distribution with mean equal to the population mean (<math display="inline"><mo>&mu;</mo></math>) and standard deviation equal to the population standard error of the mean (i.e. variance = <math display="inline"><mfrac bevelled="true"><mrow><msup><mo>&sigma;</mo><mn>2</mn></msup></mrow><mrow><mi>n</mi></mrow></mfrac></math>).

Figure 2 also shows that the mean of the sample means (`r muxbar`) is very close to the predicted population mean (`r mu`) and that the standard deviation of the sample means (`r sxbar`) is very close to the predicated standard error of the mean (`r sem`).

Note that the mean of all samples (xbar = `r xbar`) is identical to the mean of the sample means (muxbar = `r muxbar`).

The standard deviation of sample means (`r sxbar`) is significantly smaller than the population standard deviation, sigma (`r sigma`). This will be because most of the variance of the randomly sampled data is essentially discarded when we use only the means of those sampled values.

***

### Appendix

#### Introduction Code
```{r ap1, eval=FALSE, echo=TRUE}
<<ggInit>>
```
```{r ap1a, eval=FALSE, echo=TRUE}
<<intro>>
```
#### Distribution of Sample Code
```{r ap2, eval=FALSE, echo=TRUE}
<<expSampleCalc>>
```
```{r ap2a, eval=FALSE, echo=TRUE}
<<expSamplePlot>>
```
#### Distribution of Sample Means Code
```{r ap3, eval=FALSE, echo=TRUE}
<<expSampleMeansCalc>>
```
```{r ap3a, eval=FALSE, echo=TRUE}
<<expSampleMeansPlot>>
```