---
title: "Exploratory Analysis of the ToothGrowth dataset"
author: "Graham King"
date: "9 January 2015"
output: html_document
---
```{r opts, echo=FALSE}
library(knitr)
opts_chunk$set(comment=NA, fig.width=12, fig.height=6)
```
Overview
========
From the ToothGrowth help page (in datasets package):

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

```{r}
library(datasets)
data(ToothGrowth)
str(ToothGrowth)
```
From this I can see that tooth length is a continuous variable named `len`; delivery method is a factor variable named `supp` with levels `OJ` and `VC`; and the dose level is a numeric variable named `dose`. Given that dose represents three dose levels, this would be better as a factor:

```{r}
ToothGrowth$dose <- factor(ToothGrowth$dose)
print(head(ToothGrowth))

```
It should be noted that the ten subjects are not identified in the dataset.

The `supp` variable represents two delivery methods of Vitamin C, being orange juice and asorbic acide. However the levels are named 'OJ' and 'VC'. Presumably VC represents Ascorbic Acid. For clarity, I will rename it accordingly.

```{r}
levels(ToothGrowth$supp) <- c("Orange Juice", "Ascorbic Acid")
```

```{r}
library(ggplot2)
g1 <- ggplot(data=ToothGrowth, aes(x = len, fill=supp)) +
    geom_density(alpha=0.2) +
    labs(x="Length", y="Density") +
    guides(fill = guide_legend(title="Supplement")) +
    ggtitle("Figure 1: Distribution of tooth length by Supplement") +
    theme(plot.title = element_text(size = rel(1.5)))
print(g1)

oj <- subset(ToothGrowth, supp==levels(ToothGrowth$supp)[1])
vc <- subset(ToothGrowth, supp==levels(ToothGrowth$supp)[2])
g2 <- ggplot(
    data = data.frame(
        stats = matrix(c(mean(oj$len), var(oj$len), mean(vc$len), var(vc$len)), nrow=4),
        stat = factor(rep(c("mean", "variance"), times=2)),
        supp = factor(rep(levels(ToothGrowth$supp), c(2,2)))),
    aes(x=supp, y=stats, fill=stat)) +
    geom_bar(stat="identity", alpha=0.8, position="dodge") +
    labs(x="Supplement", y="Value") +
    guides(fill = guide_legend(title="Statistic")) +
    ggtitle("Figure 2: Mean and variance of tooth length by supplement") +
    theme(plot.title = element_text(size = rel(1.5)))
print(g2)
```
Figure 1 shows that tooth length is greater with the Orange Juice supplement than with the VC supplement.

Figure 2 also shows with a lower mean in the Ascorbic Acid subset, although it has a greater variance.

```{r}
g3 <- ggplot(data=ToothGrowth, aes(x = len, fill = dose)) +
    geom_density(alpha = 0.2) +
    labs(x="Tooth length", y="Density") +
    guides(fill = guide_legend(title="Dose (mg)")) +
    ggtitle("Figure 3: Distribution of tooth length by Dose") +
    theme(plot.title = element_text(size = rel(1.5)))
print(g3)

g4 <- ggplot(data=ToothGrowth, aes(x = len, fill=dose)) +
    geom_density(alpha = 0.2) +
    facet_grid(. ~ supp) +
    labs(x="Tooth length", y="Density") +
    guides(fill = guide_legend(title="Dose (mg)")) +
    ggtitle("Figure 4: Distribution of tooth length by Dose and Supplement") +
    theme(plot.title = element_text(size = rel(1.5)))
print(g4)

```